---
title: "Peer Assignment 1"
author: "Scott Artmann"
date: "April 4, 2019"
output: html_document
---

#Peer Assignment #1 - Reproducible Research
First project for the **Reproducible Research** course in Coursera's Data Science specialization track. The purpose of the project was to display both the results and the approach to solving a series of questions using data collected from a [FitBit](http://en.wikipedia.org/wiki/Fitbit).


##Synopsis
The purpose of this project was to demonstrate:

* Data loading and preprocessing
* Imputing missing values
* Answering research questions with data

## Data
Pulled from the course web site:

* Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]

The three variables included in this dataset are:

* **Steps**: Number of steps taking in a 5-minute interval (missing
    values are coded as `NA`)

* **Date**: The date on which the measurement was taken in YYYY-MM-DD
    format

* **Interval**: Identifier for the 5-minute interval in which
    measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## Loading and preprocessing the data

Download, unzip and load data into data frame `data`. 
```{r}
if(!file.exists("getdata-projectfiles-UCI HAR Dataset.zip")) {
        temp <- tempfile()
        download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
        unzip(temp)
        unlink(temp)
}
fitbitdata <- read.csv("activity.csv")
```


## What is mean total number of steps taken per day?
Sum steps by day, create Histogram, and calculate mean and median.
```{r} 
steps_by_day <- aggregate(steps ~ date, fitbitdata, sum)
hist(steps_by_day$steps, main = paste("Total Steps/Day"), col="blue", xlab="Number of Steps")
rmean <- mean(steps_by_day$steps)
rmedian <- median(steps_by_day$steps)
```

The `mean` is `r rmean` and the `median` is `r rmedian`.

## What is the average daily activity pattern?

* Calculate average steps for each interval for all days. 
* Plot the Average Number Steps per Day by Interval. 
* Find interval with most average steps. 
```{r}
steps_by_interval <- aggregate(steps ~ interval, fitbitdata, mean)
plot(steps_by_interval$interval,steps_by_interval$steps, type="l", xlab="Interval", ylab="Number of Steps",main="Average Number of Steps per Day by Interval")
max_interval <- steps_by_interval[which.max(steps_by_interval$steps),1]
```

This is the 5-minute interval, on average across all the days in the data set. Maximum number of steps is `r max_interval`.

## Impute missing values. Compare imputed to non-imputed data.
Simple imputation approach. 
Missing values were imputed by inserting the average for each interval. Thus, if interval 10 was missing on 10-02-2012, the average for that interval for all days (0.1320755), replaced the NA. 
```{r}
incomplete <- sum(!complete.cases(fitbitdata))
imputed_data <- transform(fitbitdata, steps = ifelse(is.na(fitbitdata$steps), steps_by_interval$steps[match(fitbitdata$interval, steps_by_interval$interval)], fitbitdata$steps))
```

Zeroes were forced for 10-01-2012. 
```{r}
imputed_data[as.character(imputed_data$date) == "2012-10-01", 1] <- 0
```

Recalculate total steps by day and plot Histogram. 
```{r}
steps_by_day_i <- aggregate(steps ~ date, imputed_data, sum)
hist(steps_by_day_i$steps, main = paste("Total Steps Each Day"), col="blue", xlab="Number of Steps")
#Create Histogram to show difference. 
hist(steps_by_day$steps, main = paste("Total Steps Each Day"), col="red", xlab="Number of Steps", add=T)
legend("topright", c("Imputed", "Non-imputed"), col=c("blue", "red"), lwd=10)
```

Here's the new mean and median including imputed data. 
```{r}
rmean.i <- mean(steps_by_day_i$steps)
rmedian.i <- median(steps_by_day_i$steps)
```

Here's the difference between imputed and non-imputed data.
```{r}
mean_diff <- rmean.i - rmean
med_diff <- rmedian.i - rmedian
```

Here's the total difference.
```{r}
total_diff <- sum(steps_by_day_i$steps) - sum(steps_by_day$steps)
```
* The imputed data mean is `r rmean.i`
* The imputed data median is `r rmedian.i`
* The difference between the non-imputed mean and imputed mean is `r mean_diff`
* The difference between the non-imputed mean and imputed mean is `r med_diff`
* The difference between total number of steps between imputed and non-imputed data is `r total_diff`. Thus, there were `r total_diff` more steps in the imputed data.


## Are there differences in activity patterns between weekdays and weekends?
Compared and contrasted steps during week and on weekends. There is a higher peak earlier on weekdays, and more overall activity on weekends.  
``` {r}
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", 
              "Friday")
imputed_data$dow = as.factor(ifelse(is.element(weekdays(as.Date(imputed_data$date)),weekdays), "Weekday", "Weekend"))
steps_by_interval_i <- aggregate(steps ~ interval + dow, imputed_data, mean)
library(lattice)
xyplot(steps_by_interval_i$steps ~ steps_by_interval_i$interval|steps_by_interval_i$dow, main="Average Steps per Day by Interval",xlab="Interval", ylab="Steps",layout=c(1,2), type="l")
```